---
title: "Draft post"
description: |
  An example post.
author:
  - name: Bruna Wundervald
    url: https://hamilton-institute.github.io/docs/
date: 08-27-2020
categories:
  - drafts
site: distill::distill_website  
collections:
  posts:
    citations: true
    disqus: posts
citation_url: https://hamilton-institute.github.io/docs
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 2
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a draft post for this website. This means 
that while the option `draft: true` is set in the 
preambule, the post will not show in the website, even
if the Rmarkdown file is compiled to html. When
that's changed, it will come up as a regular post. 
This option is recommended while your post still isn't
ready for publication. 

To create a draft post, just run:

```{r, eval = FALSE}
library(distill)
create_post("Name of post", draft = TRUE)

```

and add the text and other preambule options to your post. 

Other preambule options that might be useful to add to your post:

  - categories, which is here set to be "drafts", but
  could be tags such as "transport", "survival analysis", etc
  - collections, where you can define if you want
  to generate a bibtex for this post (citations: true). You
  also need to remember to set a citation URL, to which
  the citations will refer to
  - the toc options, which will set a table of contents of the post, just at the top of your post 
  

After your post is done, just run:

```{r, eval = FALSE}
library(rmarkdown)
render_site()
```


and add your changes to the GitHub repository as
indicated in the contribution guide. Happy writing!
  

